---
title: "36611 HW2"
author: "[your name here["
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Setting required files and constants

```{r constants, echo=FALSE}
## Required files ##
#   readme.txt: info from client
#   camelCase.R: (read from github) converts variable names
#                to camel case
DEMOG_FILE = "demogs.csv" # patient demographs
VISIT_FILE = "visits.csv" # patient clinic visits
DX_FILE = "DxCodes.csv" # diagnosis codes

## Other constants
DX_COL_COUNT = 3  # client says there are up to 3 diagnoses per visit

# Column names (orignal and improved)
demogColsOrig = c("ID", "Gender: 0=female; 1=male",
                 "Primary Race: 1=white; 2=black; 4=Asian",
                 "Years of Education", "Date of Birth", "Onset Age")
demogColsOrig = make.names(demogColsOrig)
demogColsImpr = c("id", "male", "race", "education", "dob", "onsetAge")

visitColDrop = make.names(c("MMS7", "MMSW", "CDR Total Score"))
visitsColsOrig = c("ID", "Date", "DX1", "DX2", "DX3",
                   "MMS", "CDR Sum of Boxes", 
                   "MisidentifyPeople", "MisidentifySelf", "Paranoid",
                   "TelevisionIsReal", "DeadPersonStillAlive",
                   "HouseNotHome")
visitsColsOrig = make.names(visitsColsOrig)
visitsColsImpr = c("id", "date", "dx1", "dx2", "dx3",
                   "MMS", "CDRBox",
                   "misidentifyPeople", "misidentifySelf", "paranoid",
                   "televisionIsReal", "deadPersonStillAlive",
                   "houseNotHome")
```



# Read in files, checking and improving column names
```{r read and fix cols, echo=FALSE, comment=NA}
demog = try(read.csv(DEMOG_FILE, as.is=TRUE), silent=TRUE)
if (is(demog, "try-error")) {
  stop("cannot read file '", DEMOG_FILE, '" in ', getwd())
}
if (ncol(demog) != length(demogColsOrig)) {
  stop("original version of ", DEMOG_FILE, " had ", length(demogColsOrig),
       " columns, but new version has ", ncol(demog))
}
demogMatch = match(demogColsOrig, names(demog))
missCol = is.na(demogMatch)
if (any(missCol)) {
  stop("missing field(s) in ", DEMOG_FILE, ": ",
   paste(demogColsOrig[missCol], collapse=", "))
}
names(demog)[demogMatch] = demogColsImpr
cat("Initial structure of demographics data.frame:\n")
str(demog)

visits = try(read.csv(VISIT_FILE, as.is=TRUE), silent=TRUE)
if (is(visits, "try-error")) {
  stop("cannot read file '", VISIT_FILE, '" in ', getwd())
}
# Drop unneeded columns
visits = visits[, names(visits) %in% visitColDrop == FALSE]
if (ncol(visits) != length(visitsColsOrig)) {
  stop("original version of ", VISIT_FILE, " had ", length(visitsColsOrig),
       " columns, but new version has ", ncol(visits))
}
visitsMatch = match(visitsColsOrig, names(visits))
missCol = is.na(visitsMatch)
if (any(missCol)) {
  stop("missing field(s) in ", VISIT_FILE, ": ",
       paste(visitsColsOrig[missCol], collapse=", "))
}
names(visits)[visitsMatch] = visitsColsImpr
rm(demogMatch, missCol, visitsMatch)
cat("\nInitial structure of visits data.frame:\n")
str(visits)
```


# Check key ("id")
```{r check key, echo=FALSE, comment=NA}
cat("Checking 'id' key\n")
dupDemog = duplicated(demog$id)
if (any(dupDemog)) {
  stop("Duplicate ids found in ", DEMOG_FILE, ": ",
       paste(demog$id[dupDemog], collapse=", "))
} else {
  cat("All ids are unique in ", DEMOG_FILE, "\n")
}

orphanDemog = setdiff(demog$id, visits$id)
if (length(orphanDemog) > 0) {
  cat("Id(s) of patients with no visits: ",
      paste(orphanDemog, collapse=", "), "\n")
  demog = demog[demog$id %in% orphanDemog == FALSE, ]
  cat("These have been removed.\n")
}

orphanVisits = setdiff(visits$id, demog$id)
if (length(orphanVisits) > 0) {
  stop("Id(s) of patients with visits but no demographics: ",
       paste(orphanVisits, collapse=", "))
}
rm(dupDemog, orphanDemog, orphanVisits)
cat("'id' key is OK\n")

# Histogram of visits
# WARNING: Always use exclude=NULL with table() in EDA
# Otherwise you will not see the missing values!
visitCounts = table(table(visits$id, exclude=NULL))
# Note: histogram() handles zeros and discrete "x" poorly
barplot(visitCounts, xlab="Number of visits",
        ylab="Number of patients", main="POP Visit Distribution")
rm(visitCounts)
```

# Fixup demog: Construct date and factors
```{r fix demog, echo=FALSE, comment=NA}
demogDateNA = is.na(demog$dob)
newDates = as.Date(demog$dob, format="%Y-%m-%d")
if (sum(is.na(newDates)) != sum(demogDateNA)) {
  badDates = !demogDateNA & is.na(newDates)
  print(demog[badDates, c("id", "dob")])
  rm(badDates)
  stop(sum(is.na(newDates)) - sum(demogDateNA),
       " demog date(s) is/are invalid.")
}
demog$dob = newDates
rm(demogDateNA, newDates)

cat("Original race distribution:")
table(demog$race, exclude=NULL)
demog$black = as.numeric(demog$race == 2)
demog$race = factor(demog$race, levels=c(1, 2, 4),
                    labels=c("white", "black", "Asian"))
cat("\nRecoded race:")
with(demog, table(race, black, exclude=NULL))

# Add a coarser education variable
cat("\nOriginal education distribution:")
table(demog$education, exclude=NULL)
demog$educ2 = demog$education
demog$educ2[demog$educ2 < 11] = 11
demog$educ2[demog$educ2 > 12 & demog$educ <16] = 15
demog$educ2[demog$educ2 > 16] = 17
demog$educ2 = factor(demog$educ2, levels=c(11, 12, 15, 16, 17),
                     labels=c("<HS", "HS", "some college",
                              "college", ">college"))
# verify correct recoding
cat("\nRecoded education distribution:")
table(demog$education, demog$educ2, exclude=NULL)

cat("\nSummary of demographics:\n")
summary(demog)
```

# Fixup visits
# Code symptoms with 8 or 9 as NA
```{r fix visits, echo=FALSE, comment=NA}
# Change "date" to a Date
visitDateNA = is.na(visits$date)
newDates = as.Date(visits$date, format="%Y-%m-%d")
if (sum(is.na(newDates)) != sum(visitDateNA)) {
  badDates = !visitDateNA & is.na(newDates)
  print(visits[badDates, c("id", "date")])
  rm(badDates)
  stop(sum(is.na(visits$date)) - nMissVisitDate,
       " visit dates are invalid.")
}
visits$date = newDates
rm(visitDateNA, newDates)

# Code NAs for dx and symptoms
for (dx in paste0("dx", 1:DX_COL_COUNT)) {
  visits[visits[, dx] == -2, dx] = NA
}

symptomCols = match(c("misidentifyPeople", "misidentifySelf",
                      "paranoid", "televisionIsReal",
                      "deadPersonStillAlive"),
                    names(visits))
cat("original symptom distributions:\n")
lapply(visits[, symptomCols], table, exclude=NULL)
# [Note: appropriate use of "for" loop]
for (col in symptomCols) {
  visits[visits[, col] %in% c(8, 9), col] = NA
}
cat("\nFinal symptom distributions:\n")
lapply(visits[, symptomCols], table, exclude=NULL)

# Create psychoisSum variable
visits$psychosisSum = apply(visits[, symptomCols], 1,
                            sum, na.rm=TRUE)

cat("\nSummary of visits:\n")
summary(visits)
```

# Univariate non-graphical EDA
```{r univariate non-graphical EDA, echo=FALSE, comment=NA}
# Provide code for these variables: race, educ2, onset,
#   first digit of dx1, first digit of dx2, CDRBox, MMS,
#   paranoid, houseNotHome, psychosisSum
# Make a few key conclusions.


```
Conclusions:




# Univariate graphical EDA
```{r univariate graphical EDA, echo=FALSE, comment=NA}
# Provide code for these variables: race, educ2, onsetAge,
#   first digit of dx1, first digit of dx2, CDRBox, MMS,
#   paranoid, houseNotHome, psychosisSum
# Make a few key conclusions.


```
Conclusions:




# Bivariate non-graphical EDA
```{r bivariate non-graphical, echo=FALSE, comment=NA}
cat("Cross-tabulation of race and education\n")
with(demog, table(race, educ2, exclude=NULL))

# Note differences in output from different coding:
cat("\nCross-tabulation of race and education")
table(demog$race, demog$educ2, exclude=NULL)

# Note how to change details of the output
cat("\nPercent at each educational level by race\n")
temp = with(demog, prop.table(table(black, educ2, exclude=NULL),
                              margin=1))
# Note: it would be easy to get black/non-black backwards
dimnames(temp) = list(race=c("non-black", "black"),
                      education=colnames(temp))
round(100*temp, digits=0)
rm(temp)


cat("\nCorrelation of education with onsetAge: ")
cat(with(demog, cor(education, onsetAge)), "\n")

cat("\nCorrelation of symptoms\n")
round(cor(visits[, symptomCols], use="pairwise.complete.obs"), digits=2)
```
Conclusions:

Years of education are weakly negatively correlated with age of onset.  This suggests that
less educated people develop Alzheimer's disease later in life.


# Bivariate graphical EDA
```{r bivariate graphical, echo=FALSE, comment=NA}
# Provide code showing the relationship between educ2 and race,
# between psychosis sum (DV) and dx1<200 (IV), and between
# psychosis sum (DV) and MMS.
# Make a few key conclusions.

```
Conclusions:

